\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

This textbook grew out of the Math Foundations course notes used to support the follow-on Data Analysis course. The project is indebted to Dr. Bassett for the course structure, mathematical direction, and original instructional framing that made these notes possible.

We also acknowledge the scribes and highlighted note takers whose work preserved the course material and made this rewrite possible:

Adrian Tomaneng, Mary Poltrack, Anna Delaney, Robert Moore, Brandon De La Rosa, Jessica Sanders, Brian Park, Duke Brady, Paul Deist, Camille Herman, Ryan Berry, Greg Lucarelli, Joshua Brewster, Christopher Posadas, Daniel Valdelamar, Katie Hernandez, Katerina Franco, Jackson Franco, Kevin Crush, Ryan Warren, Ian Moss, Keisha Castillo, Shaun Ketner, Leigh Freitag, Mariano Amaro, Samantha Brooker, and Christine La.

After the original course notes were preserved by the class scribes and note takers, Chris Posadas carried out the textbook rewrite in close collaboration with OpenAI's ChatGPT and Codex. Chris supplied the course context, student perspective, editorial judgment, and final decisions about structure, scope, wording, and mathematical emphasis. ChatGPT served as an AI drafting and review partner: helping turn the notes into textbook prose, check derivations, compare explanations against the source notes, preserve links to the Data Analysis textbook, and identify where clarity, concision, or mathematical precision could be improved. Codex supported the source-code side of the project by implementing targeted LaTeX edits, compiling test builds, checking references and labels, and reporting whether proposed changes were safe to integrate.

The workflow was deliberately iterative. Work typically moved back and forth between ChatGPT and Codex: ChatGPT helped propose or audit conceptual changes, Codex applied those changes to the source files and performed build-level quality assurance, and Chris reviewed the results before deciding whether to accept, revise, or reject the change. This process helped keep the rewrite grounded in the original Math Foundations notes while also protecting the mathematical connections to the Data Analysis textbook. Future students or instructors may be able to adapt the same human-directed, AI-assisted process: preserve the source material, map dependencies explicitly, draft in small controlled passes, audit before integrating, and use compiled builds to judge whether the result remains accurate, teachable, and coherent.

Any remaining errors in organization, exposition, notation, or interpretation belong to the textbook project, not to the original course contributors.
